---
title: "Candidate Trial Analysis Template"
author: "Lian Rui"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

## Packages

```{r packages}
library(readxl) # Import excel
library(readr) # Read text with delim "|"
library(tidyverse)
library(lubridate)
library(ggraph) # Network plot
```

```{r}
source("~/Library/Mobile Documents/com~apple~CloudDocs/Project2026/OSA and siRNA/R/load_candidate.R")
```


## Load Raw data

### Load MeSH tables

```{r MeSH setup}
# labels
mesh_labels <- read_csv("~/Documents/MeSH/mesh_labels.csv")

# parents in tree structure
mesh_parent <- read_csv("~/Documents/MeSH/mesh_parent.csv")

# tree structure
mesh_tree <- read_csv("~/Documents/MeSH/mesh_tree.csv")

# Tree name
tree_names <- read_csv("~/Documents/MeSH/tree_names.csv")
```

#### TODO: How to avoid the repeat call of `read_csv`, any elegant methods. 

### Load AACT tables

```{r AACT setup}
# studies
studies <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/studies.txt", 
                      delim = "|", escape_double = FALSE, trim_ws = TRUE)

# Select columns of studies
studies <- studies |>
  select(nct_id, 
         study_first_submitted_date, 
         start_date, 
         completion_date, 
         study_type, 
         brief_title, 
         official_title, 
         overall_status, 
         phase, 
         enrollment, 
         source, 
         source_class) |>
  filter(source_class == "INDUSTRY")


# MeSH condition
browse_conditions <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/browse_conditions.txt",
                                delim = "|", escape_double = FALSE, trim_ws = TRUE)

# conditions
# conditions <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/conditions.txt",
#                                 delim = "|", escape_double = FALSE, trim_ws = TRUE)

# MeSH intervention
browse_interventions <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/browse_interventions.txt",
                               delim = "|", escape_double = FALSE, trim_ws = TRUE)

# Interventions.txt
# interventions <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/interventions.txt",
#                                delim = "|", escape_double = FALSE, trim_ws = TRUE)

# intervention_other_names.txt
# intervention_other_names <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/intervention_other_names.txt",
#                                delim = "|", escape_double = FALSE, trim_ws = TRUE)

# Countries
# For the disease area analysis, country is not needed
countries <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/countries.txt", 
                         delim = "|", escape_double = FALSE, trim_ws = TRUE)

# design_groups
#design_groups <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/design_groups.txt", 
#                            delim = "|", escape_double = FALSE, trim_ws = TRUE)
## design_groups is not very useful for now. 

# designs outcomes
# The endpoint of each trial
# design_outcomes <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/design_outcomes.txt",
#                               delim = "|", escape_double = FALSE, trim_ws = TRUE)

# designs <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/designs.txt", 
#     delim = "|", escape_double = FALSE, trim_ws = TRUE)
# View(designs)

# facilities
#facilities <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/facilities.txt", 
#                         delim = "|", escape_double = FALSE, trim_ws = TRUE)
## Put facilities on hold

# sponsors
sponsors <- read_delim("/Volumes/My Passport/aact_Aug_2026_csv/sponsors.txt", 
                       delim = "|", escape_double = FALSE, trim_ws = TRUE)
```

#### TODO: new approach, maybe modular to avoid many `read_delim`. 

#### TODO: all the AACT files in this path `/Volumes/My Passport/aact_Aug_2026_csv/`, create a path variable and paste the file name, e.g., `sponsors.txt` would be more clean. 

### Load candidate nct_id

```{r}
path <- "~/Library/Mobile Documents/com~apple~CloudDocs/Project2026/OSA and siRNA/ASO and siRNA/"

nct_id <- "aso nct_id sep.xlsx"

nct_id_link <- paste0(path, nct_id)
```

```{r}
candidate <- load_candidate(nct_id_link)
```

#### TODO: function ready

## Candidate analysis

### Trend: Number of trials each year in last five year

```{r number of trials by year}
trial_number_year <- studies |>
  select(nct_id, study_first_submitted_date) |>
  left_join(candidate, by = join_by(nct_id)) |>
  drop_na() |>
  mutate(year = year(study_first_submitted_date)) 

dat <- trial_number_year |>
  group_by(year) |>
  summarise(n = n())

ggplot(dat) + 
  geom_col(aes(x = year, y = n), 
           fill = "steelblue") + 
  labs(title = "Trend of of ASO trials in Global", 
       y = "Number of trials") + 
  theme_minimal()
```

### Trend: Duration of trials by year, phase

```{r duration of trials by year and phase}
trial_duration <- studies |>
  select(nct_id, start_date, completion_date, phase) |>
  mutate(year = year(start_date))|>
  filter(year >= 2020) |>
  left_join(candidate, by = join_by(nct_id)) |>
  drop_na() |>
  mutate(
    time_interval = round(
      time_length(interval(start_date, completion_date), "months"),
      1
    )
  )
  
ggplot(trial_duration, aes(x = factor(year), y = time_interval)) + 
  geom_point() + 
  geom_boxplot(aes(x = factor(year))) + 
  facet_grid(phase~.)
```

### Trend: enrollment

```{r enrollment by year and phase}
enrollment <- studies |>
  select(nct_id, study_first_submitted_date, phase, enrollment) |>
  mutate(year = year(study_first_submitted_date)) |>
  filter(year >= 2021) |>
  left_join(candidate, by = join_by(nct_id)) |>
  filter(!is.na(phase))

ggplot(enrollment, aes(x = factor(year), y = enrollment)) +
  geom_point() + 
  geom_boxplot() + 
  facet_grid(phase~., scales = "free_y")
```

#### TODO Trend: enrollment by disease

### Trend: Number of trials by sponsor

```{r number of trials by sponsor}
number_trial_sponsor <- studies |>
  select(nct_id, study_first_submitted_date, phase, source) |>
  mutate(year = year(study_first_submitted_date)) |>
  filter(year >= 2021) |>
  filter(!is.na(phase)) |>
  right_join(candidate, by = join_by(nct_id)) |>
  drop_na() |>
  summarise(n = n(), .by = c(source, phase)) |>
  group_by(phase) |>
  slice_max(n, n = 10, with_ties = FALSE) |>
  arrange(phase, desc(n)) |>
  ungroup()

ggplot(number_trial_sponsor, aes(reorder(source, n, FUN = sum), n))  + 
  geom_col(aes(fill = phase)) + 
  labs(title = "Number of ASO trials by sponsor and phase", 
       x = "Sponsors", 
       y = "") + 
  coord_flip() + 
  theme_minimal()
```

### Global MeSH analysis

```{r Join candidate with aact}
# Filter the candidate studies from the master studies tables
# Create candidate specific study table. 
candidate_study <- studies %>%
  filter(nct_id %in% candidate$nct_id)

# join candidate_study with browse_condition
candidate_study <- candidate_study %>%
  left_join(browse_conditions, by = "nct_id")

# Join candidate_study with countries
candidate_study <- candidate_study %>%
  left_join(countries, by = "nct_id")

# Join candidate_study with facilities
#adc_study <- adc_study %>%
#  left_join(facilities, by = "nct_id")
## Put facilities on hold

# Join candidate_study with sponsors
candidate_study <- candidate_study %>%
  left_join(sponsors, by = "nct_id")
```

```{r rename columns}
# Rename the candidate_study
candidate_study <- candidate_study |>
  rename(country = name.x, 
         company = name.y)
```

```{r column selection}
# Select columns
candidate_study <- candidate_study %>%
  select(nct_id, study_first_submitted_date, start_date, 
         completion_date, study_type, brief_title, official_title, 
         overall_status, phase, enrollment, source, source_class, 
         downcase_mesh_term, mesh_type, country, company)

# There is a column source in study, which is the info of a study
# the source is contained in table sponsor, column company
# a single study may have many companies, e.g, 
# NCT02508090 has two companies: Santaris Pharma A/S and Hoffmann-La Roche
# As such, remove the column company, but the agency_type should be kept to
# Filter the Industry studies
candidate_study <- candidate_study |>
  select(!company)
```

```{r filter industry study}
# The industry only
candidate_study <- candidate_study %>%
  filter(source_class %in% "INDUSTRY") |>
  distinct()
```

```{r remove mesh-ancestor}
# Each study nct_id has many MeSH terms, 
# Most of them are mesh-ancestor
# i.e., the upper hierarchies of a special condition
# e.g., cystic fibrosis has ancestors like congenital, hereditary, and neonatal diseases and abnormalities, genetic diseases, inborn...
# As such, to remove the mesh-ancestor mesh type
# the more detailed hierarchy will be recreated using standalone mesh data
candidate_study <- candidate_study |>
  filter(mesh_type %in% "mesh-list") |>
  distinct()

# Join the drug name in table candidate
candidate_study <- candidate_study |>
  left_join(candidate, by = join_by(nct_id))
```

```{r network functions}
source("R/mesh_network_with_drug.R")
```

```{r network plot}
# ---------------------------------------------------------
# 1. Prepare trial data
# ---------------------------------------------------------

ta <- candidate_study |> 
  filter( phase == "PHASE1", 
          mesh_type == "mesh-list", 
          source_class == "INDUSTRY" ) |> 
  select(-any_of(c("country", "year"))) |> 
  distinct() |> 
  mutate( year = lubridate::year(study_first_submitted_date) ) |> 
  relocate( year, .after = study_first_submitted_date ) |> 
  filter(year >= 2021) |> 
  arrange(year)

# ---------------------------------------------------------
# 2. Count trials for each MeSH condition
# ---------------------------------------------------------

condition_trials <- ta |>
  distinct(
    nct_id,
    downcase_mesh_term
  ) |>
  count(
    downcase_mesh_term,
    name = "n_trials"
  ) 

# ---------------------------------------------------------
# 3. Get MeSH hierarchy
# ---------------------------------------------------------

candidate_hierarchy <- get_hierarchy(
  unique(ta$downcase_mesh_term),
  tree_names
)

candidate_hierarchy <- ta |>
  select(downcase_mesh_term, drug) |>
  left_join(
    candidate_hierarchy,
    by = c( "downcase_mesh_term" = "query_name")
  )

candidate_hierarchy <- candidate_hierarchy |>
  rename("query_name" = "downcase_mesh_term")

# ---------------------------------------------------------
# 4. Compress hierarchy
#    Cxx -> disease
#    + number of trials
# ---------------------------------------------------------

hierarchy_network <- compress_hierarchy(
  candidate_hierarchy,
  condition_trials
)

# ---------------------------------------------------------
# 5. Create network
# ---------------------------------------------------------

candidate_network <- make_mesh_network(
  hierarchy_network
)

plot_mesh_network(candidate_network)
```

## China study

```{r filter China}
# China study
candidate_china <- candidate_study %>%
  filter(country %in% "China")

# Remove the column drug
# This represent the global
# Not all the drugs are used in China
candidate_china <- candidate_china |>
  select(!drug)

# Join browse intervention by nct_id in China
candidate_china <- candidate_china |>
  mutate(year = year(study_first_submitted_date)) |>
  relocate(year, .after = study_first_submitted_date) |>
  filter(year >= 2021) |>
  left_join(browse_interventions, by = join_by(nct_id)) |>
  select(
    -mesh_term, # keep downcase column only
    -id
  ) |>
  rename("disease" = "downcase_mesh_term.x",
         "drug" = "downcase_mesh_term.y") |>
  filter(mesh_type.x == "mesh-list", 
         mesh_type.y == "mesh-list") |>
  distinct() |>
  arrange(phase)

candidate_china <- candidate_china |>
  select(nct_id,brief_title, official_title,  drug, source, phase, enrollment, disease)
```

```{r select candidate China columns}
candidate_china <- candidate_china |>
  select(nct_id, drug, source, phase, enrollment, disease) |>
  distinct()
```

```{r write table}
write_csv(candidate_china, "candidate_china.csv")
```


## TODO

1. Modular import of MeSH and AACT, no repeated `read_csv` or `read_delim`

2. `ASO` and `siRNA` as candidate category variable and change all the titles in output plot, table etc automatically

3. Explore and modify the ggplot output, auto save a plot as `PNG` with size fitted the keynote slide paste. 

4. Repeat over a phase string `"PHASE3, PHASE2, PHASE1/PHASE2, PHASE 1"`and save each network plot